Globally-centered autocovariances in MCMC

### 2020-08-25 Paper/main.tex
Layer-1 edits.

A key quantity that drives MCMC output analysis is the autocovariance function (ACvF). From their use in autocorrelation (ACF) plots to assessing Monte Carlo variability of estimators, to determining when to stop MCMC simulation, autocovariances drive many visual and quantitative inferences from MCMC output. However, estimators of ACvF are only available for single-chain implementations, with multiple-chain obtained by naive averaging. As we will demonstrate, this defeats the purpose of running parallel Markov chains from dispersed starting values.

Let $F$ be the target distribution with mean $\mu$ defined on a space $\X \subseteq \mathbb{R}^d$ equipped with a countably generated $\sigma$-field $\mathcal{B}(\X)$. For $s = 1, \dots, m$, let $\{X_{st}\}_{t\geq1}$ be the $s${th} Harris ergodic $F$-invariant Markov chain \citep[see][for definitions]{meyn:twee:2009} employed to learn characteristics about $F$. The process is covariance stationary so that
the ACvF at lag $k$ is defined as 
%
\[
    \Gamma(k) = \Cov_F(X_{s1}, X_{s\,1+k})= \mathbb{E}_F \left[(X_{s1} - \mu)(X_{s\,1+k} - \mu)^T \right]\,.
\]
Estimating $\Gamma(k)$ is critical to assessing the quality of the sampler and the reliability of Monte Carlo estimators. Let $\bar{X}_s = n^{-1} \sum_{t=1}^{n} X_{st}$ denote the Monte Carlo estimator of $\mu$ from the $s$th chain. The standard estimator for $\Gamma(k)$ is the sample autocovariance matrix at lag $k \geq 0$:
%
\begin{equation} \label{eq:empirical_ACvF}
    \hat{\Gamma}_s(k) = \dfrac{1}{n}\sum_{t=1}^{n-k} \left(X_{st} - \bar{X}_s \right) \left(X_{s\, t + k} - \bar{X}_s \right)^T\,,
\end{equation}
and for $k < 0$, $\hat{\Gamma}_s(k) = \hat{\Gamma}_s(-k)^T$. For single-chain MCMC runs, the estimator $\hat{\Gamma}_s(k)$ is used to construct ACF plots, to estimate the long-run variance of Monte Carlo estimators \cite{hannan:1970,dame:1991}, and to estimate effective sample size (ESS) \cite{kass:carlin:gelman:neal:1998,gong:fleg:2016,vats:fleg:jon:2019}. However, there is no unified approach to constructing estimators of $\Gamma(k)$ for multiple-chain implementations. Specifically, parallel MCMC chains are often spread across the state space so as to adequately capture high density areas. For slow mixing chains, the chains take time to traverse the space so that $\bar{X}_s$  are all vastly different. Consequently, $\hat{\Gamma}_s$ typically underestimates $\Gamma$ leading to a false sense of security about the quality of process.

@@

Estimators of ACvFs are often needed to estimate the long-run variance of Monte Carlo averages. Specifically, spectral variance (SV) estimators  are used to estimate $\Sigma = \sum_{k=-\infty}^{\infty} \Gamma(k)$ \citep{andr:1991,dame:1991,fleg:jone:2010}. We replace $\hat{\Gamma}(k)$ with G-ACvF in SV estimators and demonstrate strong consistency under weak conditions. In the spirit of \cite{andr:1991}, we also obtain large-sample bias and variance of the resulting estimators. SV estimators can be prohibitively slow to  compute \citep{liu:fleg:2018}. To relieve the computational burden, we adapt the fast algorithm of \cite{heberle2017fast} for G-ACvFs that dramatically reduces computation time. The globally-centered spectral variance estimator (G-SVE) is employed in the computation of ESS. We will show that using G-SVE for estimating $\Sigma$ safeguards users against early termination of the MCMC process.

@@

\caption{ACF plots using local-centering (left) and global-centering (right). Histogram estimates at $n = 10^4$ with the blue curve being estimates from $n = 50000$ {\color{red}can you do this at $10^5$?} for chain 1.}
 % (b) $n = 10^4$; the dotted lines are ACF for each chain and the thick solid line is the ACF averaged over both chains.}
\label{fig:gaussian-acf}
\end{figure}

@@

A critical use of autocovariances is in the assessment of Monte Carlo variability of estimators.  Let $g:\X \to \mathbb{R}^p$ be an $F$-integrable function so that interest is in estimating
$\mu_g = \mathbb{E}_F[g(X)]$.
% \[
% \mu_g = \mathbb{E}_F[g(X)] = \int_{\X}g(x)F(dx)\;.
% \]
 % 
Set $\{Y_{st}\}_{t \geq 1} = \{g(X_{st})\}_{t \geq 1}$ for all $s = 1, \dots, m$. Let $\bar{Y}_s = n^{-1}\sum_{t=1}^{n}Y_{st}$  and  $\bar{\bar{Y}} = m^{-1}\sum_{s=1}^{m}\bar{Y}_s$. By Birkhoff's ergodic theorem,  $\bar{\bar{Y}} \to \mu_g$ with probability 1 as $n \to \infty$. An asymptotic sampling distribution may be available via a Markov chain central limit theorem (CLT) if there exists a $p \times p$ positive-definite matrix $\Sigma$ such that
%
\begin{equation}
\label{eq:CLT}
  \sqrt{mn}(\bar{\bar{Y}} - \mu_g) \xrightarrow{d} N(0,\Sigma)\, \quad \text{ where, }
\end{equation}
% where
\begin{equation}
\label{eq:sigma}
  \Sigma = \sum_{k = -\infty}^{\infty} \Cov_F \left( Y_{11}, Y_{1(1+k)} \right) := \sum_{k = -\infty}^{\infty}\Upsilon (k)
\end{equation}
%
The goal in output analysis for MCMC is to estimate $\Sigma$ in order to assess variability in $\bar{\bar{Y}}$ \citep{fleg:hara:jone:2008,roy:2019,vats:rob:fle:jon:2020}. There is a rich literature on estimating $\Sigma$ for single-chain MCMC implementations. The most common are SV estimators \citep{andr:1991,vats:fleg:jon:2018} and batch means estimators \citep{chen:seila:1987,vats:fleg:jon:2019}. Recently, \cite{gupta:vats:2020} constructed a replicated batch means estimator for estimating $\Sigma$ from parallel chains. Batch means estimators are computationally more efficient than SV estimators, whereas SV estimators are more reliable \citep{damerdji:1995,fleg:jone:2010}. Here, we utilize G-ACvF estimators to construct globally-centered SV (G-SV) estimator of $\Sigma$. Using the methods of \cite{heberle2017fast}, we also provide an efficient algorithm for obtaining the G-SV estimator.

The locally and globally-centered  estimators of $\Upsilon(k)$ are $\hat{\Upsilon}_s(k)$ and $\hat{\Upsilon}_{G,s}(k)$, respectively, where for $k \geq 0$
\[
\hat{\Upsilon}_s(k) = \dfrac{1}{n} \ds \sum_{t=1}^{n - k} (Y_{st} - \bar{Y}_s)(Y_{s\,t+k} - \bar{Y}_s)^T \quad \text{ and } \quad \hat{\Upsilon}_{G,s}(k) = \dfrac{1}{n} \ds \sum_{t=1}^{n - k} (Y_{st} - \bar{\bar{Y}})(Y_{s\,t+k} - \bar{\bar{Y}})^T\,,
\] 
and let $\hat{\Upsilon}_G(k) = m^{-1} \sum_{k=1}^{m}\hat{\Upsilon}_{G,s}(k)$. SV estimators are constructed as weighted and truncated sums of estimated ACvFs. For some $c \geq 1$, let $w: \mathbb{R} \to [-c,c]$ be a lag window function and $b_n \in \mathbb{N}$ be a bandwidth. The (locally-centered) SV estimator of $\Sigma$ for chain $s$ is
%
%
\begin{equation} \label{eq:sve}
    \hat{\Sigma}_{s} = \sum_{k=-n+1}^{n-1}w\left(\dfrac{k}{b_n}\right)\hat{\Upsilon}_s(k)\,.
\end{equation}
Large-sample properties of $\hat{\Sigma}_s$ have been studied widely. \cite{vats:fleg:jon:2018} provided conditions for strong consistency while \cite{fleg:jone:2010,vats:flegal:2018} provided large-sample bias and variance results. These results also extend naturally to an average SV (ASV) estimator $\hat{\Sigma}_A = m^{-1} \sum_{s=1}^{m}\hat{\Sigma}_{s}$.
% \[
% \hat{\Sigma}_A = \dfrac{1}{m}\sum_{s=1}^{m}\hat{\Sigma}_{s}\;.
% \]

@@

The SV estimator, despite good statistical properties, poses limitations due to slow computation. From \eqref{eq:empirical_ACvF} and \eqref{eq:sve}, the complexity of the SV estimator is $\mathcal{O}(n^2 p^2)$ in general, and for the Bartlett lag window, it is $\mathcal{O}(b_n n p^2)$. For slow mixing Markov chains, $n$ and $b_n$ are large, limiting the use of SV estimators.  To overcome this, we adapt the  fast Fourier transform based algorithm of \cite{heberle2017fast}. Suppose $w_k = w(k/b_n)$ and let $T(w)$ be the $n \times n$ Toeplitz matrix with the first column being $(1 ~ w_1 ~ w_2 ~ \dots, ~ w_{n-1})^T$. Notice an alternate formulation of $\hat{\Sigma}_s$
%
\begin{equation} \label{eq:kyriakoulis}
    \hat{\Sigma}_s = \dfrac{1}{n}A_s^T T(w) A_s, \qquad \textrm{ where } A_s = \begin{pmatrix}
    Y_{s1} - \bar{Y}_s  & \dots & Y_{sn} - \bar{Y}_s
\end{pmatrix}^T
\end{equation}
 \cite{heberle2017fast} compute the $n \times 1$ vector $T(w)A_s$  using an algorithm based on fast Fourier transforms. Let $w^* = (1 ~ w_1 ~ w_2 ~ \dots, ~ w_{n-1}, ~0, ~w_{n-1}, \dots, w_1)^T$ and set $C(w^*)$ to be a symmetric circulant matrix such that $C(w^*)_{1:n, 1:n} = T(w)$.  With inputs $C(w^*)$ and $A_s$, Algorithm~\ref{algo:herberle} produces $\hat{\Sigma}_s$ exactly. For more details, see \cite{heberle2017fast}.
 % Let $C(w^*) = V\Lambda V^*$ be the spectral decomposition of $C(w^*)$, then $\Lambda$ is determined by a discrete Fourier transform of the first column of $C(w^*)$. 
% $A_s$ is extrapolated into a $2n \times p$ matrix $A_s^*$ by appending $n$ rows of $0$ vector to write Equation~\ref{eq:kyriakoulis} in the following form
% \[
%     \hat{\Sigma}_s = \dfrac{1}{n} A_s^T T(w) A_s = \dfrac{1}{n} A_s^T (C(w^*) A_s^*)_{1:n, :} = \dfrac{1}{n} A_s^T (V \Lambda V^* A_s^*)_{1:n, :}).
% \]
% Here $M_{1:r, \, 1:c}$ denotes the matrix truncation up to first $r$ rows and $c$ columns.

@@

\For{$j = 1, 2, \dots, p$}    
    { 
    Calculate $V^*A_s^{*_{(j)}}$ by DFT of $A^*_{s(j)}$.\;
    Multiply $V^* A_{s(j)}^{*^{(i)}}$ with the eigenvalue $\lambda_i$ for all $i = 1, ..., 2n$ to construct $\Lambda V^* A_{s(j)}^*$.\;
    Calculate $C(w^*)A^*_{s(j)} = V \Lambda V^* A_{s(j)}^*$ by inverse FFT of $\Lambda V^* A_{s(j)}^*$.\;
    }
 Select the first $n$ rows of $C(w^*)A^*_s$ to form $T(w)A_s$.\;
 Premultiply by $A_s^T$ and divide by $n$.\;
 \KwOut{$\hat{\Sigma}_s$}
 \caption{\cite{heberle2017fast} Algorithm}
 \label{algo:herberle}
\end{algorithm}

% For our purpose, we center the chain $\{Y_{st}; \; t \in \mathbb{Z}\}$ around $\bar{\bar{Y}}$ instead of $\bar{Y}_s$ in Equation \ref{eq:kyriakoulis} for all $s \in \{1, ..., m\}$. Herberle's algorithm is then applied on the formulation
We note a similar decomposition is possible for the G-SV estimator. Setting $B_s = (Y_{s1} - \bar{\bar{Y}} \; \dots \; Y_{sn} - \bar{\bar{Y}})^T$ and calling Algorithm~\ref{algo:herberle} with inputs $C(w^*)$ and $B_s$ yields $\hat{\Sigma}_{G,s}$. This algorithm has complexity $\mathcal{O}(p n \log n)$ and is thus orders of magnitude faster. Of particular importance is the fact that the bandwidth $b_n$ has close to no impact on the computation time.
% subsubsection fast_implementation (end)

@@

In this section we consider three different target distributions and sample multiple Markov chains using MCMC methods to experimentally analyse the performance of our globally-centered estimators. We will make the following three comparisons - (1) A-ACF vs G-ACF through ACrF plots; (2) A-SVE vs G-SVE; (3) $\widehat{\textrm{ESS}}_G$ vs $\widehat{\textrm{ESS}}_A$. The quality of estimation of $\Sigma$ is studied by coverage probabilities for a $95 \%$ confidence interval in cases where the true mean $\mu$ is known. The convergence of local and global estimators as $n$ increases is studied through two types of running plots (1) logarithm of Frobenius norm of estimated $\Sigma$ denoted by $\|M\|_F$ for a matrix $M$, and (2) logarithm of estimated ESS/$mn$. In all the examples we consider the function $g(x) = x$, which implies the Markov chain $\{Y_t\}$ and $\{X_t\}$ are the same.

% There are only a handful of MCMC chains where the true autocovariance and asymptotic variance are known. To compare the globally and locally-centered estimators when the truth is known, we use a slowly mixing vector autoregressive process of order 1 (VAR(1)). A distinct advantage of our estimators is observed when the target distribution is multi-modal. For this purpose we use a bivariate bi-modal probability distribution introduced by \cite{gelman1991note}. Through this example, we will also show that in case of nicely mixing Markov chains, these estimators give almost equivalent results. Lastly, we consider a real-life example of finding unknown locations of sensors using noisy distance data. The posterior distribution is this case is marginally bimodal in all dimensions. All the examples have been selected to display some kind of \textit{sticky} nature in the Markov chains. In such scenario, we successfully show that our globally-centered estimators yield better results as compared to the classical single-chain estimators for autocovariance, asymptotic variance, and effective sample size.

@@

Our first example is one where the $\Upsilon$ and $\Sigma$ are available in closed-form, so we can compare quality of estimation against a known truth. Consider a $p$-dimensional VAR(1) process $\{X_t\}_{t \geq 1}$ such that
%
\[
X_t = \Phi X_{t-1} + \epsilon_t
\]
%
where $X_t \in \mathbb{R}^p$, $\Phi $ is a $p \times p $ matrix, $ \epsilon_t \overset{i.i.d.}{\sim} N(0, \Omega)$, and $\Omega$ is a positive definite $p \times p$ matrix. We fix $\Omega$ be a AR correlation matrix with parameter .9. The invariant distribution for this Markov chain is $N(0, \Psi)$, where $vec(\Psi) = (I_{p^2} - \Phi \times \Phi)^{-1} vec(W)$. We are interested in estimating the mean of the stationary distribution, so $g$ is the identity function.  For $k > 0$, the lag-$k$ autocovariance is $\Upsilon(k) = \Gamma(k) = \Phi^k\Psi$. The process satisfies a CLT for $\bar{X} = n^{-1} \sum_{t=1}^{n} X_t$ if the spectral norm of $\Phi$ is less than 1 \citep{10.2307/1427459} and the limiting covariance, $\Sigma$ is known in closed form \citep{dai:jon:2017}. We set $p = 2$ and set $\Phi$ such that its eigenvalues are .999 and .001. The resulting process is slowly mixing and allows us to highlight the difference in the estimation of ACvF.We further set $m = 5$ with starting values dispersed across the state space.  
\begin{figure}[htbp]
\centering
   \includegraphics[width=.5\linewidth]{plots/var-acf_n1000.pdf}
     \includegraphics[width=.5\linewidth]{plots/var-acf_n10000.pdf} 
\caption{VAR(1). ACF (left) and G-ACF (right) for the second chain for $m = 5$. (Top) $n = 10^3$  and (bottom) $n =  10^4$. The red line is the true ACF.}
\label{fig:var-acf}
\end{figure}

@@

Since $\mu_g$ is known, we assess the performance of ASV and GSV by assessing coverage probabilities of $95\%$ Wald confidence regions over 1000 replications.  Table~\ref{tab:var-coverage} shows that for each sample size, the $\hat{\Sigma}_G$ results in close to nominal coverage probability, whereas $\hat{\Sigma}_A$ yields critically low coverage. The low coverage is a consequence of underestimating the autocovariances. It is only at the sample size of $n=10^5$ that $\hat{\Sigma}_A$ yields close to nominal coverage.

@@

The quality of estimation of $\Sigma$ and $\text{ESS}$ is assessed by running plots from 50 replications of run length 50000. 
 In Figure~\ref{fig:var-frob_n_ess}, we present running plots of $\log(\|\hat{\Sigma}\|_F)$ for both $\hat{\Sigma}_G$ and $\hat{\Sigma}_A$ and the running plots of $\log(\widehat{\text{ESS}})$ for both  $\widehat{\text{ESS}}_G$ and $\widehat{\text{ESS}}_A$. It is evident that the locally centered estimator of $\Sigma$ severely underestimates the truth, leading to an overestimation of ESS. The GSV estimator is able to estimate $\Sigma$ more accurately early on, safeguarding against early termination using ESS. 
\begin{figure}[htbp]
    \centering
      \includegraphics[width = 2.0in]{plots/var-frob.pdf} 
      \includegraphics[width = 2.0in]{plots/var-ess.pdf}         
    \caption{VAR(1). (Left) Running plot for logarithm of Frobenius norm of A-SV and G-SV.(Right) Running plot for logarithm of $\widehat{\textrm{ESS}}/mn$ using A-SV and G-SV.}
    \label{fig:var-frob_n_ess}
\end{figure}

>>>>

A key quantity that drives MCMC output analysis is the autocovariance function (ACvF). From their use in autocorrelation (ACF) plots to assessing Monte Carlo variability of estimators, to determining when to stop MCMC simulation, autocovariances drive many visual and quantitative inferences from MCMC output. However, estimators of ACvF are only available for single-chain implementations, with multiple-chain version obtained by naive averaging. As we will demonstrate, this defeats the purpose of running parallel Markov chains from dispersed starting values.

Let $F$ be the target distribution with mean $\mu$ defined on a space $\X \subseteq \mathbb{R}^d$ equipped with a countably generated $\sigma$-field $\mathcal{B}(\X)$. For $s = 1, \dots, m$, let $\{X_{st}\}_{t\geq1}$ be the $s${th} Harris ergodic $F$-invariant Markov chain \citep[see][for definitions]{meyn:twee:2009} employed to learn characteristics about $F$. The process is covariance stationary so that
the ACvF at lag $k$ is defined as 
%
\[
    \Gamma(k) = \Cov_F(X_{s1}, X_{s\,1+k})= \mathbb{E}_F \left[(X_{s1} - \mu)(X_{s\,1+k} - \mu)^T \right]\,.
\]
Estimating $\Gamma(k)$ is critical to assessing the quality of the sampler and the reliability of Monte Carlo estimators. Let $\bar{X}_s = n^{-1} \sum_{t=1}^{n} X_{st}$ denote the Monte Carlo estimator of $\mu$ from the $s$th chain. The standard estimator for $\Gamma(k)$ is the sample autocovariance matrix at lag $k \geq 0$:
%
\begin{equation} \label{eq:empirical_ACvF}
    \hat{\Gamma}_s(k) = \dfrac{1}{n}\sum_{t=1}^{n-k} \left(X_{st} - \bar{X}_s \right) \left(X_{s\, t + k} - \bar{X}_s \right)^T\,,
\end{equation}
and for $k < 0$, $\hat{\Gamma}_s(k) = \hat{\Gamma}_s(-k)^T$. For single-chain MCMC runs, the estimator $\hat{\Gamma}_s(k)$ is used to construct ACF plots, to estimate the long-run variance of Monte Carlo estimators \citep{hannan:1970,dame:1991}, and to estimate effective sample size (ESS) \citep{kass:carlin:gelman:neal:1998,gong:fleg:2016,vats:fleg:jon:2019}. However, there is no unified approach to constructing estimators of $\Gamma(k)$ for multiple-chain implementations. Specifically, parallel MCMC chains are often spread across the state space so as to adequately capture high density areas. For slow mixing chains, the chains take time to traverse the space so that $\bar{X}_s$  are all vastly different. Consequently, $\hat{\Gamma}_s$ typically underestimates $\Gamma$ leading to a false sense of security about the quality of process.

@@

Estimators of ACvFs are often needed to estimate the long-run variance of Monte Carlo averages. Specifically, spectral variance (SV) estimators  are used to estimate $\Sigma = \sum_{k=-\infty}^{\infty} \Gamma(k)$ \citep{andr:1991,dame:1991,fleg:jone:2010}. We replace $\hat{\Gamma}(k)$ with G-ACvF in SV estimators to obtain globally-centered SV (G-SV) estimator and demonstrate strong consistency under weak conditions. In the spirit of \cite{andr:1991}, we also obtain large-sample bias and variance of the resulting estimator. SV estimators can be prohibitively slow to  compute \citep{liu:fleg:2018}. To relieve the computational burden, we adapt the fast algorithm of \cite{heberle2017fast} for G-SV estimator that dramatically reduces computation time. The G-SV estimator  is employed in the computation of ESS. We will show that using G-SV estimator for estimating $\Sigma$ safeguards users against early termination of the MCMC process.

@@

\caption{ACF plots using local-centering (left) and global-centering (right). Histogram estimates at $n = 10^4$ with the blue curve being estimates from $n = 10^5$ for chain 1.}
 % (b) $n = 10^4$; the dotted lines are ACF for each chain and the thick solid line is the ACF averaged over both chains.}
\label{fig:gaussian-acf}
\end{figure}

@@

A critical use of autocovariances is in the assessment of Monte Carlo variability of estimators.  Let $g:\X \to \mathbb{R}^p$ be an $F$-integrable function so that interest is in estimating
$\mu_g = \mathbb{E}_F[g(X)]$.
% \[
% \mu_g = \mathbb{E}_F[g(X)] = \int_{\X}g(x)F(dx)\;.
% \]
 % 
Set $\{Y_{st}\}_{t \geq 1} = \{g(X_{st})\}_{t \geq 1}$ for all $s = 1, \dots, m$. Let $\bar{Y}_s = n^{-1}\sum_{t=1}^{n}Y_{st}$  and  $\bar{\bar{Y}} = m^{-1}\sum_{s=1}^{m}\bar{Y}_s$. By Birkhoff's ergodic theorem,  $\bar{\bar{Y}} \to \mu_g$ with probability 1 as $n \to \infty$. An asymptotic sampling distribution may be available via a Markov chain central limit theorem (CLT) if there exists a $p \times p$ positive-definite matrix $\Sigma$ such that
%
\begin{equation}
\label{eq:CLT}
  \sqrt{mn}(\bar{\bar{Y}} - \mu_g) \xrightarrow{d} N(0,\Sigma)\, \quad \text{ where, }
\end{equation}
% where
\begin{equation}
\label{eq:sigma}
  \Sigma = \sum_{k = -\infty}^{\infty} \Cov_F \left( Y_{11}, Y_{1(1+k)} \right) := \sum_{k = -\infty}^{\infty}\Upsilon (k)
\end{equation}
%
The goal in output analysis for MCMC is to estimate $\Sigma$ in order to assess variability in $\bar{\bar{Y}}$ \citep{fleg:hara:jone:2008,roy:2019,vats:rob:fle:jon:2020}. There is a rich literature on estimating $\Sigma$ for single-chain MCMC implementations. The most common are SV estimators \citep{andr:1991,vats:fleg:jon:2018} and batch means estimators \citep{chen:seila:1987,vats:fleg:jon:2019}. Recently, \cite{gupta:vats:2020} constructed a replicated batch means estimator for estimating $\Sigma$ from parallel chains. Batch means estimators are computationally more efficient than SV estimators, whereas SV estimators are more reliable \citep{damerdji:1995,fleg:jone:2010}. Here, we utilize G-ACvF estimators to construct globally-centered SV (G-SV) estimator of $\Sigma$. Using the method of \cite{heberle2017fast}, we also provide a computationally efficient algorithm for obtaining the G-SV estimator.

The locally and globally-centered  estimators of $\Upsilon(k)$ are $\hat{\Upsilon}_s(k)$ and $\hat{\Upsilon}_{G,s}(k)$, respectively, where for $k \geq 0$
\[
\hat{\Upsilon}_s(k) = \dfrac{1}{n} \ds \sum_{t=1}^{n - k} (Y_{st} - \bar{Y}_s)(Y_{s\,t+k} - \bar{Y}_s)^T \quad \text{ and } \quad \hat{\Upsilon}_{G,s}(k) = \dfrac{1}{n} \ds \sum_{t=1}^{n - k} (Y_{st} - \bar{\bar{Y}})(Y_{s\,t+k} - \bar{\bar{Y}})^T\,,
\] 
and let $\hat{\Upsilon}_G(k) = m^{-1} \sum_{k=1}^{m}\hat{\Upsilon}_{G,s}(k)$. SV estimators are constructed as weighted and truncated sums of estimated ACvFs. For some $c \geq 1$, let $w: \mathbb{R} \to [-c,c]$ be a lag window function and $b_n \in \mathbb{N}$ be a bandwidth. The (locally-centered) SV estimator of $\Sigma$ for chain $s$ is
%
%
\begin{equation} \label{eq:sve}
    \hat{\Sigma}_{s} = \sum_{k=-n+1}^{n-1}w\left(\dfrac{k}{b_n}\right)\hat{\Upsilon}_s(k)\,.
\end{equation}
Large-sample properties of $\hat{\Sigma}_s$ have been studied widely. \cite{vats:fleg:jon:2018} provided conditions for strong consistency while \cite{fleg:jone:2010,vats:flegal:2018} provided large-sample bias and variance results. These results also extend naturally to an average SV (ASV) estimator defined as $\hat{\Sigma}_A := m^{-1} \sum_{s=1}^{m}\hat{\Sigma}_{s}$.
% \[
% \hat{\Sigma}_A = \dfrac{1}{m}\sum_{s=1}^{m}\hat{\Sigma}_{s}\;.
% \]

@@

The SV estimator, despite having good statistical properties, poses limitations due to slow computation. From \eqref{eq:empirical_ACvF} and \eqref{eq:sve}, the complexity of the SV estimator is $\mathcal{O}(n^2 p^2)$ in general, and for the Bartlett lag window, it is $\mathcal{O}(b_n n p^2)$. For slow mixing Markov chains, $n$ and $b_n$ are large, limiting the use of SV estimators.  To overcome this, we adapt the  fast Fourier transform based algorithm of \cite{heberle2017fast}. Suppose $w_k = w(k/b_n)$ and let $T(w)$ be the $n \times n$ Toeplitz matrix with the first column being $(1 ~ w_1 ~ w_2 ~ \dots, ~ w_{n-1})^T$. Notice an alternate formulation of $\hat{\Sigma}_s$
%
\begin{equation} \label{eq:kyriakoulis}
    \hat{\Sigma}_s = \dfrac{1}{n}A_s^T T(w) A_s, \qquad \textrm{ where } A_s = \begin{pmatrix}
    Y_{s1} - \bar{Y}_s  & \dots & Y_{sn} - \bar{Y}_s
\end{pmatrix}^T
\end{equation}
 \cite{heberle2017fast} compute the $n \times 1$ vector $T(w)A_s$  using an algorithm based on fast Fourier transforms. Let $w^* = (1 ~ w_1 ~ w_2 ~ \dots, ~ w_{n-1}, ~0, ~w_{n-1}, \dots, w_1)^T$ and set $C(w^*)$ to be a symmetric circulant matrix such that the matrix truncation $C(w^*)_{1:n, 1:n} = T(w)$.  With inputs $C(w^*)$ and $A_s$, Algorithm~\ref{algo:herberle} produces $\hat{\Sigma}_s$ exactly. For more details, see \cite{heberle2017fast}.
 % Let $C(w^*) = V\Lambda V^*$ be the spectral decomposition of $C(w^*)$, then $\Lambda$ is determined by a discrete Fourier transform of the first column of $C(w^*)$. 
% $A_s$ is extrapolated into a $2n \times p$ matrix $A_s^*$ by appending $n$ rows of $0$ vector to write Equation~\ref{eq:kyriakoulis} in the following form
% \[
%     \hat{\Sigma}_s = \dfrac{1}{n} A_s^T T(w) A_s = \dfrac{1}{n} A_s^T (C(w^*) A_s^*)_{1:n, :} = \dfrac{1}{n} A_s^T (V \Lambda V^* A_s^*)_{1:n, :}).
% \]
% Here $M_{1:r, \, 1:c}$ denotes the matrix truncation up to first $r$ rows and $c$ columns.

@@

\For{$j = 1, 2, \dots, p$}    
    { 
    Calculate $V^*A^*_{s(j)}$ by DFT of $A^*_{s(j)}$.\;
    Multiply $V^* A_{s(j)}^{*^{(i)}}$ with the eigenvalue $\lambda_i$ for all $i = 1, ..., 2n$ to construct $\Lambda V^* A_{s(j)}^*$.\;
    Calculate $C(w^*)A^*_{s(j)} = V \Lambda V^* A_{s(j)}^*$ by inverse FFT of $\Lambda V^* A_{s(j)}^*$.\;
    }
 Select the first $n$ rows of $C(w^*)A^*_s$ to form $T(w)A_s$.\;
 Premultiply by $A_s^T$ and divide by $n$.\;
 \KwOut{$\hat{\Sigma}_s$}
 \caption{\cite{heberle2017fast} Algorithm}
 \label{algo:herberle}
\end{algorithm}

% For our purpose, we center the chain $\{Y_{st}; \; t \in \mathbb{Z}\}$ around $\bar{\bar{Y}}$ instead of $\bar{Y}_s$ in Equation \ref{eq:kyriakoulis} for all $s \in \{1, ..., m\}$. Herberle's algorithm is then applied on the formulation
We observe that a similar decomposition is possible for the G-SV estimator. Setting $B_s = (Y_{s1} - \bar{\bar{Y}} \; \dots \; Y_{sn} - \bar{\bar{Y}})^T$ and calling Algorithm~\ref{algo:herberle} with inputs $C(w^*)$ and $B_s$ yields $\hat{\Sigma}_{G,s}$. This algorithm has complexity $\mathcal{O}(p n \log n)$ and is thus orders of magnitude faster. Of particular importance is the fact that the bandwidth $b_n$ has close to no impact on the computation time.
% subsubsection fast_implementation (end)

@@

In this section we consider three different target distributions and sample $m$ parallel Markov chains using MCMC methods to experimentally analyse the performance of our globally-centered estimators. We will make the following three comparisons - (1) A-ACF vs G-ACF through ACrF plots; (2) A-SVE vs G-SVE; (3) $\widehat{\textrm{ESS}}_G$ vs $\widehat{\textrm{ESS}}_A$. The quality of estimation of $\Sigma$ is studied by coverage probabilities of a $95 \%$ Wald  confidence interval in cases where the true mean $\mu$ is known. The convergence of local and global estimators of $\Sigma$ and $\widehat{\textrm{ESS}}$ as $n$ increases is studied through two types of running plots (1) logarithm of Frobenius norm of estimated $\Sigma$, and (2) logarithm of estimated ESS/$mn$. In all the examples we are interested in estimating the mean of the stationary distribution $\mu$, so $g$ is the identity function. The batch size $b_n$ is obtained from the function \texttt{batchSize} from \texttt{R} package \texttt{mcmcse}.

% There are only a handful of MCMC chains where the true autocovariance and asymptotic variance are known. To compare the globally and locally-centered estimators when the truth is known, we use a slowly mixing vector autoregressive process of order 1 (VAR(1)). A distinct advantage of our estimators is observed when the target distribution is multi-modal. For this purpose we use a bivariate bi-modal probability distribution introduced by \cite{gelman1991note}. Through this example, we will also show that in case of nicely mixing Markov chains, these estimat give almost equivalent results. Lastly, we consider a real-life example of finding unknown locations of sensors using noisy distance data. The posterior distribution is this case is marginally bimodal in all dimensions. All the examples have been selected to display some kind of \textit{sticky} nature in the Markov chains. In such scenario, we successfully show that our globally-centered estimators yield better results as compared to the classical single-chain estimators for autocovariance, asymptotic variance, and effective sample size.

@@

Our first example is one where the $\Upsilon$ and $\Sigma$ are available in closed-form, so we can compare quality of estimation against a known truth. Consider a $p$-dimensional VAR(1) process $\{X_t\}_{t \geq 1}$ such that
%
\[
X_t = \Phi X_{t-1} + \epsilon_t
\]
%
where $X_t \in \mathbb{R}^p$, $\Phi $ is a $p \times p $ matrix, $ \epsilon_t \overset{i.i.d.}{\sim} N(0, \Omega)$, and $\Omega$ is a positive definite $p \times p$ matrix. We fix $\Omega$ be a AR correlation matrix with parameter .9. The invariant distribution for this Markov chain is $N(0, \Psi)$, where $vec(\Psi) = (I_{p^2} - \Phi \times \Phi)^{-1} vec(W)$.  For $k > 0$, the lag-$k$ autocovariance is $\Upsilon(k) = \Gamma(k) = \Phi^k\Psi$. The process satisfies a CLT for $\bar{X} = n^{-1} \sum_{t=1}^{n} X_t$ if the spectral norm of $\Phi$ is less than 1 \citep{10.2307/1427459} and the limiting covariance, $\Sigma$ is known in closed form \citep{dai:jon:2017}. We set $p = 2$ and set $\Phi$ such that its eigenvalues are .999 and .001. The resulting process is slowly mixing and allows us to highlight the difference in the estimation of ACvF.We further set $m = 5$ with starting values dispersed across the state space.  
\begin{figure}[htbp]
\centering
   \includegraphics[width=.5\linewidth]{plots/var-acf_n1000.pdf}
     \includegraphics[width=.5\linewidth]{plots/var-acf_n10000.pdf} 
\caption{VAR(1). ACF (left) and G-ACF (right) for the second chain for $m = 5$. (Top) $n = 10^3$  and (bottom) $n =  10^4$. The red line is the true ACF.}
\label{fig:var-acf}
\end{figure}

@@

Since $\mu$ is known, we assess the performance of A-SV and G-SV estimatiors by assessing coverage probabilities of $95\%$ Wald confidence regions over 1000 replications.  Table~\ref{tab:var-coverage} shows that for each sample size, the $\hat{\Sigma}_G$ results in close to nominal coverage probability, whereas $\hat{\Sigma}_A$ yields critically low coverage. The low coverage is a consequence of underestimating the autocovariances. It is only at the sample size of $n=10^5$ that $\hat{\Sigma}_A$ yields close to nominal coverage.

@@

The quality of estimation of $\Sigma$ and $\text{ESS}$ is assessed by running plots from 50 replications of run length 50000. 
 In Figure~\ref{fig:var-frob_n_ess}, we present running plots of $\log(\|\hat{\Sigma}\|_F)$ for both $\hat{\Sigma}_G$ and $\hat{\Sigma}_A$ and the running plots of $\log(\widehat{\text{ESS}})$ for both  $\widehat{\text{ESS}}_G$ and $\widehat{\text{ESS}}_A$. It is evident that the locally centered estimator of $\Sigma$ severely underestimates the truth, leading to an overestimation of ESS. The G-SV estimator is able to estimate $\Sigma$ more accurately early on, safeguarding against early termination using ESS. 
\begin{figure}[htbp]
    \centering
      \includegraphics[width = 2.0in]{plots/var-frob.pdf} 
      \includegraphics[width = 2.0in]{plots/var-ess.pdf}         
    \caption{VAR(1). (Left) Running plot for logarithm of Frobenius norm of A-SV and G-SV.(Right) Running plot for logarithm of $\widehat{\textrm{ESS}}/mn$ using A-SV and G-SV.}
    \label{fig:var-frob_n_ess}
\end{figure}
